---
title: "README"
output: md_document
---

R functions for working with syntactic structure coded as token lists (e.g. CONLL format)

Installation
====

You can install directly from github:

```{r, eval=F}
library(devtools)
install_github("vanatteveldt/rsyntax")
```

Usage
====

The functions in this module assume that you have a list of tokens in a data frame.
A simple example is provided with the module:

```{r, message=F, eval=F}
library(rsyntax)
data(example_tokens)
tokens
```

```{r, echo=F, message=F}
library(rsyntax)
data(example_tokens)
library(knitr)
kable(tokens)
```

Get the text of a sentence, optionally specifying which column(s) to use:

```{r}
get_text(tokens)
get_text(tokens, word.column = c("lemma", "pos"))
```

Plot the syntactic structure of a sentence:
(Note: if you have multiple sentences in one token list, you should filter it or provide a sentence= argument)

```{r readme_example_plot, fig.path=".", fig.cap="Syntactic Structure of example sentence"}
g = graph_from_sentence(tokens)
plot(g)
```

Clauses and Sources
====

You can use the `get_quotes` function to extract quotes and paraphrases from the sentences.
Note that for this, the token ids need to be globally unique. 
If that is not the case, you can use the `unique.ids` function to make them unique:

```{r}
tokens = unique_ids(tokens)
```

You can get the quotes from the tokens with `get_quotes`:

```{r, eval=F}
quotes = get_quotes(tokens)
quotes
```
```{r, echo=F}
quotes = get_quotes(tokens)
kable(quotes)
```

A single quote was found, with node 2 ("say") as the key, node 1 ("John") as the  sources, and nodes 3 through 6 ("that Mary hit him") as quote. 

To find the clauses, you can use the get_clauses function, which takes the quotes as an optional argument to make sure that speech actions are not listed as clauses:

```{r, eval=F}
clauses = get_clauses(tokens, quotes=quotes)
clauses
```
```{r, echo=F}
clauses = get_clauses(tokens, quotes=quotes)
kable(clauses)
```

Finally, you can also provide the quotes and clauses to the `graph_from_sentence` function. 
This will fill the clauses in a desaturated rainbow, with the subject as a circle and the predicate as rectangle.
Quotes are represented with a bright node for the source, and the border in the same colour for the quote.

```{r readme_example_plot_clauses, fig.path=".", fig.cap="Syntactic Structure of example sentence with clauses and quotes marked"}
g = graph_from_sentence(tokens, quotes = quotes, clauses = clauses)
plot(g)
```
