---
title: "README"
output: md_document
---

This repository contains the package `amcatr`,
which has a number of useful functions for working with the AmCAT API in R.

# Installing


You can install `amcatr` directly from this github repository uising devtools:

```{r}
if (!require(devtools)) {install.packages("devtools"); library(devtools)}
install_github("amcat/amcat-r")
```

# Connecting

Use the `amcat_connect()` function to connect to an AmCAT server. The typical mode of use is to 

```{r}
?amcat_co 
library(amcatr)
amcat.save.password("https://amcat.nl", username="example", password="secret")
```

Note: You only need to run this line once per computer, so *do not save this line* in a script, and certainly do not publish or email it, to avoid compromising the security of your account!

Now, you can connect to AmCAT using `amcat.connect`.
This requests an authentication token from the specified AmCAT server and stores it for further commands.
The resulting data (named `conn`, below) is then used every time you call the API. 

```r
library(amcatr)
conn = amcat.connect("https://amcat.nl")
```

Usage
-----

All functions in `amcatr` start with `amcat.` and you can use the built-in help functions to view documentation. 

To help you get started with the AmCAT API, please see the following how-to guides:

+ [Howto: Querying](howto/howto_query.md) deals with querying and metadata retrieval. 
+ [Howto: DTM](howto/howto_create_document_term_matrix.md) shows how you can create a document term matrix, and refers to usefull packages for computational text analysis. 
