---
title: "README"
output: md_document
---

R functions for working with syntactic structure coded as token lists (e.g. CONLL format)

Installation
====

You can install directly from github:

```{r, eval=F}
library(devtools)
install_github("kasperwelbers/corpus-tools")
```

Usage
====

The functions in this module assume that you have a list of tokens in a data frame.
A simple example is provided with the module:

```{r, message=F, eval=F}
library(rsyntax)
data(example_tokens)
head(tokens)
```

```{r, echo=F, message=F}
library(rsyntax)
data(example_tokens)
library(knitr)
kable(head(tokens))
```

Get the text of a sentence, optionally specifying which column(s) to use:

```{r}
get_text(tokens)
get_text(tokens, word.column = c("lemma", "pos"))
```

Plot the syntactic structure of a sentence:
(Note: if you have multiple sentences in one token list, you should filter it or provide a sentence= argument)

```{r readme_example_plot, fig.path=".", fig.cap="Syntactic Structure of example sentence"}
g = graph_from_sentence(tokens)
plot(g)
```